# Command Compendium

This chapter is designed to help you easily craft commands.

[ggplot color options](colors%20are%20here:%20https://sape.inf.usi.ch/quick-reference/ggplot2/colour)

## Basic Data Manipulation

### Creating Objects

Creating an object allows you to store the output of a command in the environment. This means you can reuse the result later by referencing the object name, instead of rerunning the command every time.

```{r, eval=F}

## When this command runs, it will show up in the environment
chart <- ggplot(DATASET_OBJECT, aes(y = VARIABLE OF INTEREST)) +
  geom_boxplot(fill = "lightgrey")

##now I can "call" the object
chart

```

### Filtering Data

To create our analytical dataset we often need to subset a larger dataset. Filtering allows us to simplify. We always want to save filtered data as objects so that we can reference them in other commands.

```{r, eval=F}
#to select certain columns
columns <- DATASET |> select(COLUMN1, COLUMN2, COLUMN3)

### IF VALUES ARE STRING (NON-NUMERIC) THEY MUST BE IN "" 

#to filter by values in a certain column
values <- DATASET |> filter(COLUMN == VALUE)

# to filter by multiple conditions. 
#and
values <- DATASET |> filter(COLUMN == VALUE & COLUMN2 == VALUE2)

#or 
values <- DATASET |> filter(COLUMN == VALUE | COLUMN2 == VALUE2)

#in 
values <- DATASET |> filter(COLUMN %in% c(VALUE1, VALUE2, VALUE3))

#numeric operator
values <- DATASET |> filter(COLUMN < 2000)

```

### Making New Columns

```{r, eval=F}
#creating new columns
dataset <- DATASET |> mutate(NEWCOLUMN = COMMAND, NEWCOLUMN2 = COMMAND)

```

### Renaming Columns

```{r, eval=F}
#renaming columns
clean_data <- DATASET |> rename(
  new_name1 = old_name1,
  new_name2 = old_name2
)

```

### Grouping and Summarizing Data

```{r, eval=F}

summary_data <- DATASET |> 
  group_by(CATEGORY_COLUMN) |> 
  summarize(
    mean_value = mean(NUMERIC_COLUMN, na.rm = TRUE),
    count = n()
  )
```

## Descriptive Statistics

### Descriptive Statistics Tables

```{r, eval=F}
#requires e1071 package
#requires gt package
## descriptive statistics for quantitative data
  DATASET_OBJECT |> #whatever the object is named
  st_drop_geometry() |> #gets rid of geometry column
  select(VARIABLE_OF_INTEREST) |> #field name of variable of interest
  summarise(                  #summarizes variable of interest. You can add/remove
    n = n(),
    num_na = sum(is.na(VARIABLE_OF_INTEREST)), 
    mean = mean(VARIABLE_OF_INTEREST, na.rm = TRUE),
    median = median(VARIABLE_OF_INTEREST, na.rm = TRUE),
    sd = sd(VARIABLE_OF_INTEREST, na.rm = TRUE),
    variance = var(VARIABLE_OF_INTEREST, na.rm = TRUE),
    mean_dev = mean(abs(VARIABLE_OF_INTEREST - mean(VARIABLE_OF_INTEREST, na.rm = TRUE)), na.rm = TRUE),
    cv = sd / mean,
    min = min(VARIABLE_OF_INTEREST, na.rm = TRUE),
    max = max(VARIABLE_OF_INTEREST, na.rm = TRUE),
    skewness = skewness(VARIABLE_OF_INTEREST, na.rm = TRUE), 
    kurtosis = kurtosis(VARIABLE_OF_INTEREST, na.rm = TRUE)
                               
  ) |> 
    #this starts the table formatting.
  pivot_longer(everything(), names_to = "Statistic", values_to = "Value") |>
  gt() |>
  tab_header(
    title = "TITLE", ##Add your title here
  ) %>%
  fmt_number(
    columns = everything(),
    decimals = 2 ##You can change number of decimals
  )

## Descriptive statistics table for categorical data
DATASET_OBJECT |> #whatever the object is named
  st_drop_geometry() |> #gets rid of geometry column
  group_by(ruca_code) |> #variable that you want to summarize
  summarise(n = n())  |> #calculates the count
  gt() #formats into a table

```

### Histogram

```{r, eval=F}
#requires ggplot2 package
## Create a basic histogram. 
## You will likely need to modify the binwidth command
ggplot(DATASET_OBJECT, aes(x = VARIABLE OF INTEREST)) +
  geom_histogram(binwidth = 1, fill = "lightgrey", color = "black", alpha = 0.7) +
  labs(
    title = "TITLE",
    x = "X-AXIS LABEL",
    y = "Y-AXIS LABEL"
  ) 

```

### Boxplot

```{r, eval = F}
#requires ggplot2 package
# create basic boxplot
ggplot(DATASET_OBJECT, aes(y = VARIABLE OF INTEREST)) +
  geom_boxplot(fill = "lightgrey") +
  labs(
    title = "TITLE",
    y = "Y-AXIS LABEL"
  )

```

### Violin Plot

```{r,  eval = F}
#requires ggplot2 package
##you will likely need to modify the bw
ggplot(DATASET_OBJECT, aes(x = 1, y = VARIABLE_OF_INTEREST)) +
  geom_violin(fill = "lightgrey", bw = 1.2)  + geom_boxplot(width=0.1) +
  labs(
    title = "TITLE",
    y = "Y AXIS LABEL"
  ) + theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

```

### Bar Chart

```{r, eval = F}
#requires ggplot2 package
##YOU NEED TO FIRST GROUP THE DATA
grouped_data <- DATASET_OBJECT |> st_drop_geometry() |> 
  group_by(VARIABLE_OF_INTEREST) |>
  summarise(Count = n()) 

#this plots the bar plot (note that the coord_flip swtiches the axis which is optional
ggplot(grouped_data, aes(x=VARIABLE_OF_INTEREST, y=Count)) + 
  geom_bar(stat = "identity") + coord_flip()  + labs(
    x = "X_AXIS_LABEL",
    y = "Y_AXIS_LABEL",
  )

```

## Maps

```{r, eval = F}
#required tmap package
## classification styles
## "equal"
## "fisher"
## "pretty"
## "quantile"

#command to see color palettes
cols4all::c4a_gui()

##make a basic point map
tm_shape(DATABASE_OBJECT) + tm_dots(fill = "VARIABLE_OF_INTEREST",
                fill.scale = tm_scale_intervals(values = "brewer.COLORPALETTE", style = "CLASSIFICATION", n= NUMBEROFCLASSES)) 

##make a basic polygon map
tm_shape(DATABASE_OBJECT) + tm_polygon(fill = "VARIABLE_OF_INTEREST",
                fill.scale = tm_scale_intervals(values = "brewer.COLOR_PALETTE", style = "CLASSIFICATION", n= NUMBEROFCLASSES)) 


## make an interactive map
tmap_mode(mode = "view")

##go back to static map 
tmap_mode(mode = "plot")


## add title 
MAPCOMMANDS + tm_title("TITLE")


## Add transparency
tm_shape(DATABASE_OBJECT) + tm_polygon(fill = "VARIABLE_OF_INTEREST",fill_alpha = VALUE) 

```

## Descriptive Spatial Statistics

### Mean Center and Weighted Mean Center

```{r, eval = F}
#requires sfdep package

#calculate mean center
DATASET_mean_center <- center_mean(DATASET)

#calculate weighted mean center
DATASET_mean_center <- center_mean(DATASET, weight = DATASET$VARIABLE)


```

### Standard Deviational Ellipse

```{r, eval = F}
#requires sfdep package
#calculate standard ellipse values
std_ellip_DATASET <- std_dev_ellipse(DATASET)

#create an ellipse of those values
std_ellip_DATASET <- sfdep::st_ellipse(geometry =std_ellip_DATASET,
                                   sx = std_ellip_DATASET$sx,
                                   sy = std_ellip_DATASET$sy,
                                   rotation = -std_ellip_DATASET$theta)
```
